### 2012-11-29 branches/CUDA/BookGPU/Chapters/chapter14/ch14.tex
New fig 3 and summary
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1823 40876441-3496-0410-a613-a2fca35b2ae2

\chapter{Ludwig: GPU implementation of a complex fluid lattice Boltzmann
application}

@@

\section{Introduction}
The lattice Boltzmann (LB) method (for an overview see, e.g.,
\cite{succi-book}) has become a popular approach to a variety of fluid
dynamics problems.  It provides a way to solve the incompressible,
isothermal Navier-Stokes equations and has the attractive features of
being both explicit in time and local in space. This makes the LB
method well-suited to parallel computation. Many efficient parallel
implementations of the LB method have been undertaken, typically using
a combination of distributed domain decomposition and the Message
Passing Interface (MPI). However, the potential
performance benefits offered by GPUs has motivated a new `mixed-mode'
approach to address very large problems. Here, fine-grained
parallelism is implemented on the GPU, while MPI is reserved for
larger-scale parallelism.  This mixed mode is of increasing interest
to application programmers at a time when many supercomputing services
are moving
toward clusters of GPU accelerated nodes. The design questions which
arise when developing a lattice Boltzmann code for this type of
heterogeneous system are therefore worth studying. However, similar
questions also recur in many other types of stencil-based algorithms.

The first applications of LB on GPUs were to achieve fluid-like
effects in computer animation, rather than scientific applications per
se.  These early works include simple fluids~\cite{wei2004}, miscible
two-component flow~\cite{zhu2006}, and various image processing tasks
based on the use of partial differential equations~\cite{zhao2007}.
While these early works used relatively low level graphics APIs, the
first CUDA runtime interface implementation was a two-dimensional
simple fluid problem~\cite{toelke2010}.  Following pioneering work on
clusters of GPUs coupled via MPI to study air
pollution~\cite{fan2004}, more recent work has included mixed OpenMP
and CUDA~\cite{myre2011}, Posix threads and CUDA~\cite{obrecht2011},
and MPI and CUDA for increasingly large GPU clusters
\cite{bernaschi2010,xian2011,feichtinger2011}. The heterogeneous
nature of these systems has also spurred interest in approaches
including automatic code generation \cite{walshsaar2012} and autotuning
\cite{williams2011} to aid application performance.

Many of these authors make use of
another attractive feature of LB: the ability to include fixed
solid-fluid boundary conditions as a straightforward addition to
the algorithm to study, for example, flow in
porous media. This points to an important
application area for LB methods: that of complex fluids.
Complex fluids include mixtures, surfactants, liquid crystals,
and particle suspensions, and typically require additional physics
beyond the bare Navier-Stokes equations to provide a full
description~\cite{aidun2010}. The representation of this extra
physics raises additional design questions for the application
programmer. Here, we consider the \textit{Ludwig} code \cite{desplat},
an LB application developed specifically for complex fluids
(\textit{Ludwig} was named for Boltzmann, 1844--1906).
We will present the steps
required to allow Ludwig to exploit efficiently both a single
GPU, and also many GPUs in parallel.  We show that
Ludwig scales excellently to at least the one thousand GPU level
(the largest resource available at the time of writing) with
indications that the code will scale to much larger systems as
they become available. In addition, we consider the steps required
to represent fully-resolved moving solid particles (usually referred
to as colloids in the context of complex fluids). Such particles
need to have their surface resolved on the lattice scale in order
to include relevant surface physics, and must be able to move: e.g.,
to execute Brownian motion in response to random forces from the
fluid. Standard methods are available to represent such particles
(e.g., \cite{ladd1994,nguyen2002}) which are amenable to effective
domain decomposition and message passing \cite{stratford2008}.
We describe below the additional considerations which arise
for such moving particles when developing an
implementation on a GPU.

In the following section we provide a brief overview of the lattice
Boltzmann method, and mention some of the general issues which can
influence performance in the CPU context.
In Section \ref{ch14:sec:singlegpu}, we then describe the alterations
which are required to exploit the GPU architecture effectively, and
highlight how the resulting code differs from the CPU version. In
Section \ref{ch14:sec:parallelgpu}, we extend this description to
include the steps required to allow exploitation of many GPUs in
parallel while retaining effective scaling. We also present results for
a typical benchmark for a fluid-only problem in Section
\ref{ch14:sec:parallelgpu} to demonstrate the success of the approach.
In Section \ref{ch14:sec:particles}, we describe the design choices
which are relevant to a GPU implementation of moving particles. Finally,
we include in the summary a number of general observations on software
engineering and maintenance which arise from our experience.

@@

More specifically, we store a vector of 19 double precision floating
point values at each lattice site for the distribution function
$f_i(\mathbf{r};t)$.
The collision operation ${\cal L}_{ij}$, which is local at each lattice
site, may be thought of as follows. A matrix-vector multiplication
${\cal M}_{ij}f_j$ is used to transform the distributions into the
hydrodynamic quantities, where ${\cal M}_{ij}$ is a constant 19x19
matrix related to the choice of
$\mathbf{c}_i$. The non-conserved hydrodynamic quantities are then
relaxed toward their (known) equilibrium values, and are transformed
back to new post-collision distributions via the inverse transformation
${\cal M}^{-1}_{ij}$. This gives rise to the need for a minimum of 2x19$^2$
floating point multiplications per lattice site. In contrast, the
propagation stage consists solely of moving the distribution values
between lattice sites, and involves no floating point operations.

In the CPU version, the \textit{Ludwig} implementation stores one time
level of distribution values $f_i(\mathbf{r}; t)$. This distribution
is stored as a flat array of C data type \texttt{double}, and laid out
so that all elements of the velocity are contiguous at a given site
(often referred to as ``array-of-structures'' order). This is
appropriate for sums over the distribution required locally at the
collision stage as illustrated schematically in
Listing~\ref{ch14:listing1}: the fact that consecutive loads are from
consecutive memory addresses allows the prefetcher to fully engage.
(The temporary scalar \texttt{a\_tmp} allows caching of the
intermediate accumulated value in the innermost loop.)  A variety of
standard sequential optimisations are relevant for the collision stage
(loop unrolling, inclusion of SIMD operations, and so on
\cite{wellein2006}).  For example, the collision stage in
\textit{Ludwig} includes explicit SIMD code, which is useful if the
compiler on a given platform cannot identify it.  The propagation
stage is separate, and is organised as a ``pull'' (again, various
optimisation have been considered, e.g.,
\cite{pohl2003,mattila2007,wittmann2012}).  No further optimisation is
done here beyond ensuring that the ordering of the discrete velocities
allows memory access to be as efficient as possible. While these
optimisations are important, it should be remembered that for some
complex fluid problems, the hydrodynamics embodied in the LB
calculation is a relatively small part of the total computational cost
(at the 10\%-level in some cases).  This means optimisation effort may
be better concentrated elsewhere.

@@

In this section we describe the steps taken to enable \textit{Ludwig}
for the GPU. There are a number of crucial issues: first, the
minimisation of data traffic between host and device; second, the
optimal mapping of available parallelism onto the architecure and
third, the issue of memory coalescing. We discuss each of these in
turn.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=10cm]{Chapters/chapter14/figures/two_graphs_ud}
\caption{The weak (top) and strong (bottom) scaling of Ludwig.  Closed
  shapes denote results using the CPU version run on the Cray
  XE6 (using two 16-core AMD Interlagos CPUs per node), while open
  shapes denote results using the GPU version on the Cray XK6 (using a
  single NVIDIA X2090 GPU per node). Top: the benchmark time is shown
  where the problem size per node is constant. Bottom: performance is
  shown where, for each of the four problem sizes presented, the
  results are scaled by the lattice volume, and all results are
  normalized by the same arbitrary constant.  }
\label{ch14:fig:scaling}
\end{figure}

@@

Figure \ref{ch14:fig:scaling} shows the results of both weak and
strong scaling tests (top and bottom panels, respectively). For
weak scaling, where the local sub-domain size is kept fixed
(here a relatively large 196$^3$ lattice), the time taken by
an ideal code would remain constant. It can be seen that while
scaling of the CPU version shows a pronounced upward slope, the
GPU version scales almost perfectly. The advantage of the GPU
version over the CPU version is the fact that the GPU version
has a factor of 32 fewer MPI tasks per node, so communications
require a signifcantly smaller number of larger data transfers.
The performance
advantage of the GPU version ranges from a factor of around 1.5 to
around 1.8. Careful study by other authors \cite{williams2011} have
found the absolute performance (in terms of floating point
operations per second, or floating point operations per Watt) to
be remarkably similar between architectures.

Perhaps of more interest is the strong scaling picture (lower
panel in Figure \ref{ch14:fig:scaling}) where the performance
as a function of the number of nodes is measured for fixed problem
size. We consider four different fixed problem sizes on both CPU
(up to 1024 nodes) and GPU (up to 768 nodes). To allow comparison,
the results are scaled by total system size in each case. For strong
scaling, the disparity in the number of MPI tasks is clearly revealed
in the failure of the CPU version to provide any significant benefit
beyond a modest number of nodes as communication overheads dominate.
In contrast, the GPU version shows
reasonably robust scaling even for the smaller system sizes and good
scaling for the larger systems.

@@

\begin{figure}[t]
\centering
\includegraphics[width=10cm]{Chapters/chapter14/figures/bbl}
\caption{A 2D illustration of colloidal particles (grey circles)
  interacting with the lattice, where each lattice point is
  represented by a cross. Left: each particle spans multiple sites.
  Right: the particle-facing distribution components are moved inside
  the particle with the direction reversed, such that the ``bounce
  back'' will be completed when the fluid propagates.}
\label{ch14:fig:bbl}
\end{figure}

@@

We have implemented the second option as follows. For each sub-domain,
a list of boundary-cutting links is assembled on the CPU which includes
the identity of the relevant element of the distribution. This list,
together with the particle information required to compute the correct
bounce-back term, are transferred to the GPU. The updates to the relevant
elements of the distribution can then take place on the GPU. The
corresponding information to compute the update of the particle dynamics
is returned
to the CPU, where the reduction over the surface links is computed.
The change of particle shape may be dealt with in a similar manner:
the relatively small number of updates required at any one time step
(or however frequently the particle position is updated) can be
marshaled to the GPU as necessary.
CAN WE SAY SOMETHING LIKE: This preliminary implementation
is found to be effective on problems involving up to X per cent solid
volume fraction.

Finally, we note that the CPU version actually avoids the collision calculation
at solid lattice points by consulting a look-up table of solid/fluid
status. On the GPU, it is perhaps preferable to perform the collision
stage everywhere instead of moving the look-up table to the GPU and
introducing the associated logic.
Ultimately, the GPU might favour other boundary methods which treat solid and
fluid on a somewhat more equal basis, for example, the immersed boundary
method \cite{ch14:immersed,ch14:immersed-lb} or smoothed profile method
\cite{ch14:spm}.
However, the approach adopted here  allows us to exploit
the GPU for the intensive fluid simulation whilst maintaining the complex
code required for particles on the CPU. Overheads of CPU-GPU transfer are
minimised by transferring only those data relevant to the hydrodynamic
interaction implemented via bounce-back on links.

@@

\section{Summary}

The \textit{Ludwig} LB application, which specializes
in complex fluid problems, has undergone the developments required to
use large-scale parallel GPU accelerated supercomputers effectively.
The new functionality augments the original code, so that
\textit{Ludwig} can run on either CPU-based or GPU-accelerated systems. 
Using the NVIDIA CUDA programming language, all the computational
kernels were offloaded to the GPU. This included not only those
responsible for the majority of the compute time, but any that
accessed the relevant data structures. This allows data to be kept
resident on the GPU and avoids expensive data transfers. We described
steps taken to optimise the performance on the GPU architecture by
reducing off-chip memory accesses and restructuring the data layout
in memory to ensure the available memory bandwidth was exploited fully.

A new halo-exchange communication phase for the code was developed to
allow efficient parallel scaling to many GPUs. The CPU version
relies on MPI datatype functionality for CPU to CPU communication; we
have explicitly written buffer packing/unpacking and transfer
operations to for GPU to GPU data exchange, staged through the host
CPUs. This includes the filtering of velocity components to only
transfer those required in a specific direction, hence reducing the
overall volume of data transferred. We used CUDA stream functionality
to overlap separate stages within the communication phase, and reduce
the overall communication time. The effect was measured to be more
profound for smaller local system sizes, i.e. it is especially useful
in aiding strong scaling.

We described work to enable colloidal particles in the simulation with
minimal overhead. We implemented these in such a way that we avoid a
major diversion of the CPU and GPU codebases, whilst minimising data
transfers between the CPU and GPU at each timestep. We keep the
majority of the (relatively inexpensive) particle related code on the
CPU, while offloading only those parts responsible for the interaction
with the fluid to the GPU (where the fluid data resides).

Our resulting package is able to exploit the largest of GPU
accelerated architectures to address a wide range of complex problems.

%\section{Introduction}
%% no \IEEEPARstart
%This demo file is intended to serve as a ``starter file''
%for IEEE conference papers produced under \LaTeX\ using
%IEEEtran.cls version 1.7 and later.
%% You must have at least 2 lines in the paragraph with the drop letter
%% (should never be an issue)
%I wish you the best of success.

%\hfill mds

%\hfill January 11, 2007

%\subsection{Subsection Heading Here}
%Subsection text here.

%\subsubsection{Subsubsection Heading Here}
%Subsubsection text here.

@@

\bibitem{stratford2008}
K. Stratford and I. Pagonabarraga,
Parallel domain decomposition for lattice Boltzmann with moving particles,
CAMWA REF \textbf{55}, 1585 (2008).

>>>>

\chapter{Ludwig: multiple GPUs for a complex fluid lattice Boltzmann
application}

@@

\section{Introduction}
The lattice Boltzmann (LB) method (for an overview see, e.g.,
\cite{succi-book}) has become a popular approach to a variety of fluid
dynamics problems.  It provides a way to solve the incompressible,
isothermal Navier-Stokes equations and has the attractive features of
being both explicit in time and local in space. This makes the LB
method well-suited to parallel computation. Many efficient parallel
implementations of the LB method have been undertaken, typically using
a combination of distributed domain decomposition and the Message
Passing Interface (MPI). However, the potential
performance benefits offered by GPUs has motivated a new `mixed-mode'
approach to address very large problems. Here, fine-grained
parallelism is implemented on the GPU, while MPI is reserved for
larger-scale parallelism.  This mixed mode is of increasing interest
to application programmers at a time when many supercomputing services
are moving
toward clusters of GPU accelerated nodes. The design questions which
arise when developing a lattice Boltzmann code for this type of
heterogeneous system are therefore worth studying. Furthermore, similar
questions also recur in many other types of stencil-based algorithms.

The first applications of LB on GPUs were to achieve fluid-like
effects in computer animation, rather than scientific applications per
se.  These early works include simple fluids~\cite{wei2004}, miscible
two-component flow~\cite{zhu2006}, and various image processing tasks
based on the use of partial differential equations~\cite{zhao2007}.
While these early works used relatively low level graphics APIs, the
first CUDA runtime interface implementation was a two-dimensional
simple fluid problem~\cite{toelke2010}.  Following pioneering work on
clusters of GPUs coupled via MPI to study air
pollution~\cite{fan2004}, more recent work has included mixed OpenMP
and CUDA~\cite{myre2011}, Posix threads and CUDA~\cite{obrecht2011},
and MPI and CUDA for increasingly large GPU clusters
\cite{bernaschi2010,xian2011,feichtinger2011}. The heterogeneous
nature of these systems has also spurred interest in approaches
including automatic code generation \cite{walshsaar2012} and auto-tuning
\cite{williams2011} to aid application performance.

Many of these authors make use of
another attractive feature of LB: the ability to include fixed
solid-fluid boundary conditions as a straightforward addition to
the algorithm to study, for example, flow in
porous media. This points to an important
application area for LB methods: that of complex fluids.
Complex fluids include mixtures, surfactants, liquid crystals,
and particle suspensions, and typically require additional physics
beyond the bare Navier-Stokes equations to provide a full
description~\cite{aidun2010}. The representation of this extra
physics raises additional design questions for the application
programmer. Here, we consider the \textit{Ludwig} code \cite{desplat},
an LB application developed specifically for complex fluids
(\textit{Ludwig} was named for Boltzmann, 1844--1906).
We will present the steps
required to allow \textit{Ludwig} to exploit efficiently both a single
GPU, and also many GPUs in parallel.  We show that
\textit{Ludwig} scales excellently to at least the one thousand GPU level
(the largest resource available at the time of writing) with
indications that the code will scale to much larger systems as
they become available. In addition, we consider the steps required
to represent fully-resolved moving solid particles (usually referred
to as colloids in the context of complex fluids). Such particles
need to have their surface resolved on the lattice scale in order
to include relevant surface physics, and must be able to move: e.g.,
to execute Brownian motion in response to random forces from the
fluid. Standard methods are available to represent such particles
(e.g., \cite{ladd1994,nguyen2002}) which are amenable to effective
domain decomposition and message passing \cite{stratford2008}.
We describe below the additional considerations which arise
for such moving particles when developing an
implementation on a GPU.

In the following section we provide a brief overview of the lattice
Boltzmann method, and mention some of the general issues which can
influence performance in the CPU context.
In Section \ref{ch14:sec:singlegpu}, we then describe the alterations
which are required to exploit the GPU architecture effectively, and
highlight how the resulting code differs from the CPU version. In
Section \ref{ch14:sec:parallelgpu}, we extend this description to
include the steps required to allow exploitation of many GPUs in
parallel while retaining effective scaling. We also present results for
a typical benchmark for a fluid-only problem in Section
\ref{ch14:sec:parallelgpu} to demonstrate the success of the approach.
In Section \ref{ch14:sec:particles}, we describe the design choices
which are relevant to a GPU implementation of moving particles. Finally,
we include a number of general observations on software
engineering and maintenance which arise from our experience.
A summary is provided in Section~\ref{ch14:sec:summary}.

@@

More specifically, we store a vector of 19 double precision floating
point values at each lattice site for the distribution function
$f_i(\mathbf{r};t)$.
The collision operation ${\cal L}_{ij}$, which is local at each lattice
site, may be thought of as follows. A matrix-vector multiplication
${\cal M}_{ij}f_j$ is used to transform the distributions into the
hydrodynamic quantities, where ${\cal M}_{ij}$ is a constant 19x19
matrix related to the choice of
$\mathbf{c}_i$. The non-conserved hydrodynamic quantities are then
relaxed toward their (known) equilibrium values, and are transformed
back to new post-collision distributions via the inverse transformation
${\cal M}^{-1}_{ij}$. This gives rise to the need for a minimum of 2x19$^2$
floating point multiplications per lattice site. (Additional operations are
required to implement, for example, the force $\mathbf{f}(\mathbf{r})$.)
In contrast, the
propagation stage consists solely of moving the distribution values
between lattice sites, and involves no floating point operations.

In the CPU version, the \textit{Ludwig} implementation stores one time
level of distribution values $f_i(\mathbf{r}; t)$. This distribution
is stored as a flat array of C data type \texttt{double}, and laid out
so that all elements of the velocity are contiguous at a given site
(often referred to as ``array-of-structures'' order). This is
appropriate for sums over the distribution required locally at the
collision stage as illustrated schematically in
Listing~\ref{ch14:listing1}: the fact that consecutive loads are from
consecutive memory addresses allows the prefetcher to engage fully.
(The temporary scalar \texttt{a\_tmp} allows caching of the
intermediate accumulated value in the innermost loop.)  A variety of
standard sequential optimisations are relevant for the collision stage
(loop unrolling, inclusion of SIMD operations, and so on
\cite{wellein2006}).  For example, the collision stage in
\textit{Ludwig} includes explicit SIMD code, which is useful if the
compiler on a given platform cannot identify it.  The propagation
stage is separate, and is organised as a ``pull'' (again, various
optimisation have been considered, e.g.,
\cite{pohl2003,mattila2007,wittmann2012}).  No further optimisation is
done here beyond ensuring that the ordering of the discrete velocities
allows memory access to be as efficient as possible. While these
optimisations are important, it should be remembered that for some
complex fluid problems, the hydrodynamics embodied in the LB
calculation is a relatively small part of the total computational cost
(at the 10\%-level in some cases).  This means optimisation effort may
be better concentrated elsewhere.

@@

In this section we describe the steps taken to enable \textit{Ludwig}
for the GPU. There are a number of crucial issues: first, the
minimisation of data traffic between host and device; second, the
optimal mapping of available parallelism onto the architecture and
third, the issue of memory coalescing. We discuss each of these in
turn.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=10cm]{Chapters/chapter14/figures/two_graphs_ud}
\caption{The weak (top) and strong (bottom) scaling of \textit{Ludwig}.  Closed
  shapes denote results using the CPU version run on the Cray
  XE6 (using two 16-core AMD Interlagos CPUs per node), while open
  shapes denote results using the GPU version on the Cray XK6 (using a
  single NVIDIA X2090 GPU per node). Top: the benchmark time is shown
  where the problem size per node is constant. Bottom: performance is
  shown where, for each of the four problem sizes presented, the
  results are scaled by the lattice volume, and all results are
  normalized by the same arbitrary constant.  }
\label{ch14:fig:scaling}
\end{figure}

@@

Figure \ref{ch14:fig:scaling} shows the results of both weak and
strong scaling tests (top and bottom panels, respectively). For
weak scaling, where the local sub-domain size is kept fixed
(here a relatively large 196$^3$ lattice), the time taken by
an ideal code would remain constant. It can be seen that while
scaling of the CPU version shows a pronounced upward slope, the
GPU version scales almost perfectly. The advantage of the GPU
version over the CPU version is the fact that the GPU version
has a factor of 32 fewer MPI tasks per node, so communications
require a significantly smaller number of larger data transfers.
The performance
advantage of the GPU version ranges from a factor of around 1.5 to
around 1.8. Careful study by other authors \cite{williams2011} have
found the absolute performance (in terms of floating point
operations per second, or floating point operations per Watt) to
be remarkably similar between architectures.

Perhaps of more interest is the strong scaling picture (lower
panel in Figure \ref{ch14:fig:scaling}) where the performance
as a function of the number of nodes is measured for fixed problem
size. We consider four different fixed problem sizes on both CPU
(up to 512 nodes shown) and GPU (up to 768 nodes). To allow comparison,
the results are scaled by total system size in each case. For strong
scaling, the disparity in the number of MPI tasks is clearly revealed
in the failure of the CPU version to provide any significant benefit
beyond a modest number of nodes as communication overheads dominate.
In contrast, the GPU version shows
reasonably robust scaling even for the smaller system sizes and good
scaling for the larger systems.

@@

\begin{figure}[t]
\centering
%\includegraphics[width=10cm]{Chapters/chapter14/figures/bbl}
\includegraphics[width=10cm]{Chapters/chapter14/figures/colloid_new}
\caption{
A two-dimensional schematic picture of spherical particles on the lattice.
Left: a particle is allowed
to move continuously across the lattice, and the position of the
surface defines fluid lattice sites (light blue) and solid lattice
sites (dark red). The discrete surface is defined by links
where propagation would intersect the surface (arrows). Note the
discrete shape of the two particles is different. Right: post-collision
distributions are reversed at the surface by the process of bounce-back
on links, which replaces the propagation.
%A 2D illustration of colloidal particles (grey circles)
%  interacting with the lattice, where each lattice point is
%  represented by a cross. Left: each particle spans multiple sites.
%  Right: the particle-facing distribution components are moved inside
%  the particle with the direction reversed, such that the ``bounce
%  back'' will be completed when the fluid propagates.
}
\label{ch14:fig:bbl}
\end{figure}

@@

We have implemented the second option as follows. For each sub-domain,
a list of boundary-cutting links is assembled on the CPU which includes
the identity of the relevant element of the distribution. This list,
together with the particle information required to compute the correct
bounce-back term, are transferred to the GPU. The updates to the relevant
elements of the distribution can then take place on the GPU. The
corresponding information to compute the update of the particle dynamics
is returned
to the CPU, where the reduction over the surface links is computed.
The change of particle shape may be dealt with in a similar manner:
the relatively small number of updates required at any one time step
(or however frequently the particle position is updated) can be
marshaled to the GPU as necessary. This preliminary implementation
is found to be effective on problems involving up to 4\%
solid volume fraction.

We note that the CPU version actually avoids the collision calculation
at solid lattice points by consulting a look-up table of solid/fluid
status. On the GPU, it is perhaps preferable to perform the collision
stage everywhere instead of moving the look-up table to the GPU and
introducing the associated logic.
Ultimately, the GPU might favour other boundary methods which treat solid and
fluid on a somewhat more equal basis, for example, the immersed boundary
method \cite{ch14:immersed,ch14:immersed-lb} or smoothed profile method
\cite{ch14:spm}.
However, the approach adopted here  allows us to exploit
the GPU for the intensive fluid simulation whilst maintaining the complex
code required for particles on the CPU. Overheads of CPU-GPU transfer are
minimised by transferring only those data relevant to the hydrodynamic
interaction implemented via bounce-back on links.

@@

\section{Summary}
\label{ch14:sec:summary}

We have described the steps take to implement, on both single and
multiple GPUs, the \textit{Ludwig} code which was originally
designed for complex fluid problems on a conventional CPU
architecture. We have added the necessary functionality using
NVIDIA CUDA and discussed the important changes to the main
data structures. By retaining domain decomposition and message
passing via MPI, we have demonstrated it is possible to scale
complex fluid problems to large numbers of GPUs in parallel.
By following the key design criteria of maximising parallelism
and minimising host-device data transfers, we have confirmed
the mixed MPI-GPU approach is viable for scientific applications.
For the more intricate problem of moving solid particles, we find
it is possible to retain the more serial elements related to
particle link operations on the CPU, while offloading only the
parallel lattice-based operations involving the LB distribution
to the GPU. Again, this minimises host-device movement of data.

From the software engineering viewpoint, some duplication of code
to allow efficient implementation on both host and device is
currently required. This issue might be addressed by approaches
such as automatic kernel generation, but may also be addressed
naturally in time as GPU and CPU hardware converge. Portable
abstractions and APIs, perhaps based on approaches such as OpenCL,
will also facilitate the development and maintenance of portable
codes which also exhibit portable performance (perhaps in conjunction
with automatic tuning approaches).

So, while the challenges in designing portable and efficient scientific
applications remain very real, this work provides some hope the large
clusters of GPU machines can be used effectively for a wide range of
complex fluid problems.

%The \textit{Ludwig} LB application, which specializes
%in complex fluid problems, has undergone the developments required to
%use large-scale parallel GPU accelerated supercomputers effectively.
%The new functionality augments the original code, so that
%\textit{Ludwig} can run on either CPU-based or GPU-accelerated systems. 
%Using the NVIDIA CUDA programming language, all the computational
%kernels were offloaded to the GPU. This included not only those
%responsible for the majority of the compute time, but any that
%accessed the relevant data structures. This allows data to be kept
%resident on the GPU and avoids expensive data transfers. We described
%steps taken to optimise the performance on the GPU architecture by
%reducing off-chip memory accesses and restructuring the data layout
%in memory to ensure the available memory bandwidth was exploited fully.

%A new halo-exchange communication phase for the code was developed to
%allow efficient parallel scaling to many GPUs. The CPU version
%relies on MPI datatype functionality for CPU to CPU communication; we
%have explicitly written buffer packing/unpacking and transfer
%operations to for GPU to GPU data exchange, staged through the host
%CPUs. This includes the filtering of velocity components to only
%transfer those required in a specific direction, hence reducing the
%overall volume of data transferred. We used CUDA stream functionality
%to overlap separate stages within the communication phase, and reduce
%the overall communication time. The effect was measured to be more
%profound for smaller local system sizes, i.e. it is especially useful
%in aiding strong scaling.

%We described work to enable colloidal particles in the simulation with
%minimal overhead. We implemented these in such a way that we avoid a
%major diversion of the CPU and GPU codebases, whilst minimising data
%transfers between the CPU and GPU at each timestep. We keep the
%majority of the (relatively inexpensive) particle related code on the
%CPU, while offloading only those parts responsible for the interaction
%with the fluid to the GPU (where the fluid data resides).

%Our resulting package is able to exploit the largest of GPU
%accelerated architectures to address a wide range of complex problems.

@@

\bibitem{stratford2008}
K. Stratford and I. Pagonabarraga,
Parallel domain decomposition for lattice Boltzmann with moving particles,
\textit{Comput. Math. with Applications} \textbf{55}, 1585 (2008).
